Predicting FOMC Actions using NLP; Assesing the impact of matrix sparsity and regularization

### 2019-05-14 presentation/paperv3.tex
paper
edits and adding plot for paper-v3

\section{Objectives}
\begin{enumerate}
\item Given the proliferation of robust open-source ML and NLP packages, we were driven to understand the characteristics of the feature matrix output (i.e., document-term matrix) from the text processing functions (i.e., CountVectorizer) analyzing the number of features, matrix sparsity, and matrix norm generated from utilizing packages functions.
\item Assess the effeiveness of Multinomial Niave Bayes and Logistic Lasso applied to problems where the feature size is orders of magnitues larger than the label size ($p>>n$).
\item Determine whether ML and NLP methods can be used to predict FOMC action (no change or a change in the base rate) using FOMC statements, minutes, and Board of Governors speeches.
\item Measure model performance on F1 scores because $70\%$ of the actions over the nineteen year period we are measuring are ``no action''.  F1 scores ensure that the accuracy of the model is not based purely on a naive ``no action'' guess, which in the best case would give a $70\%$ model accuracy.
\end{enumerate}

@@

\noindent \begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline
 \multicolumn{5}{|c|}{CountVector Matrix Size} \\
 \hline
 \multicolumn{1}{|c|}{} &
 \multicolumn{2}{|c|}{Unstacked} &
 \multicolumn{2}{|c|}{Stacked}\\
 \hline
 N-gram & sparsity & size(billions) & sparsity & size(billions)\\
 \hline
       1:1& 0.03628801& 0.008561&  0.10635940&   0.001757 \\
       1:2& 0.00544312& 0.170058&  0.02081416&    0.036107\\
       1:3& 0.00368457& 0.439898&  0.01508941&     0.092869\\
       3:5& 0.00238128& 0.900721&  0.01059293&     0.173040\\
       4:7& 0.00224699& 1.332560&  0.00996789&     0.271117\\
      5:10& 0.00218455& 1.995502&  0.00972276&     0.402041\\
      8:10& 0.00213633& 1.019811&  0.00955894&     0.198627\\
     10:15& 0.00210643& 2.097187&  0.00945586&     0.398312\\
     20:20& 0.00205055& 0.347043&  0.00931157&     0.074799\\
 \hline
\end{tabular}

@@

\subsubsection{A comparision of Multinomial Naive Bayes to Logistic Regression with extrem sparse Matricies}
Under extreme levels of sparseness (sparseness measure of less than $0.0021$) from  Matricies 
generated by CountVectorizer , Logistic Regression and Logistic Lasso break down while
while Multinomial Naive Bayes can still produce usable results. Logistic Lasso with relaitvely low penatly 
$\alpha=0.002$ functions as well or better than Naive Bayes when sparseness exceeded $0.0025$.
Logistic Lasso with a larger penalty $> 2.0$ does not performe will with extremley sparse matricies. \\

\noindent \begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline

@@

 \hline
 \multicolumn{1}{|c|}{N-gram} &
 \multicolumn{1}{|c|}{sparsity} &
 \multicolumn{1}{|c|}{smooth$=0.0$} &
 \multicolumn{1}{|c|}{$\alpha=0.002$} &
 \multicolumn{1}{|c|}{$\alpha=1.000$} \\
 \hline
 60& 0.001979 & 0.647 & 0.463 & 0.256\\
 50& 0.002018 & 0.619 & 0.565 & 0.342\\
 30& 0.002038 & 0.724 & 0.620 & 0.460\\
 25& 0.002063 & 0.720 & 0.664 & 0.490\\
 20& 0.002067 & 0.725 & 0.678 & 0.520\\
 12& 0.002069 & 0.682 & 0.743 & 0.656\\
 \hline
\end{tabular}

\section{Conclusion}
Reducing the sparsity of the matricies generated from CountVectorizer by stacking the documents greatly improved the results.
We were able to attain maximum test accuracy of 95.9$\%$ using a 10:15 n-gram with a recall of 93.4$\%$ and an F1 score of 92.5$\%$.
Because of the number of features generated can be in the millions, Logistic Lasso with small penatly proved to be the most effective 
method for reducing the number of coefficients.
The fraction of non-zero coefficients remaining after applying logistic lasso
ranged between 3e-05 to 1.6e-03.  (Compared to .97 with no regularization using logistic regression).
Logistic regression  and Logistic Lasso both broke down at the most extreme levels of sparsity $<0.0020$ 
In general Regularization parameters at the extremes ($\alpha=0.0$ or $\alpha > 2.0$) produced inferior results.
MultiNomial Naive Bayes with Laplace smoothing$=0$ was the most stable method for producing useful results at the
most extreme measures of matrix sparcity.  In general no smoothing or large smoothing proved to give the best results with
MultiNomial Naive Bayes. Overall Logistic Lasso with $\alpha = 0.002$ produced the most stable results.

>>>>

\section{Objectives}
\begin{enumerate}
\item Given the proliferation of robust open-source ML and NLP packages, we were driven to understand the characteristics of the feature matrix output (i.e., document-term matrix) from the text processing functions (i.e., CountVectorizer) analyzing the number of features, matrix sparsity, and matrix norm generated from utilizing packages functions.
\item Assess the effeiveness of Multinomial Niave Bayes and Logistic Lasso applied to problems where the feature size is orders of magnitues larger than the label size ($p>>n$).
\item Determine whether ML and NLP methods can be used to predict FOMC action (no change or a change in the base rate) using FOMC statements, minutes, and Board of Governors speeches.
\item Measure model performance on F1 scores because $70\%$ of the actions over the nineteen year period we are measuring are ``no action''.  F1 scores ensure that the accuracy of the model is not based purely on a naive ``no action'' guess, which in the best case would give a $70\%$ model accuracy.
\item Create scraping and cleaning tools to assemble a corpus of documents containing Speeches, Statements, Minutes and Actions from the Federal Reserve.
\end{enumerate}

@@

\noindent 
\begin{center}
\begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline
 \multicolumn{5}{|c|}{CountVector Matrix Size} \\
 \hline
 \multicolumn{1}{|c|}{} &
 \multicolumn{2}{|c|}{Unstacked} &
 \multicolumn{2}{|c|}{Stacked}\\
 \hline
 N-gram & sparsity & size(billions) & sparsity & size(billions)\\
 \hline
       1:1& 0.03628801& 0.008561&  0.10635940&   0.001757 \\
       1:2& 0.00544312& 0.170058&  0.02081416&    0.036107\\
       1:3& 0.00368457& 0.439898&  0.01508941&     0.092869\\
       3:5& 0.00238128& 0.900721&  0.01059293&     0.173040\\
       4:7& 0.00224699& 1.332560&  0.00996789&     0.271117\\
      5:10& 0.00218455& 1.995502&  0.00972276&     0.402041\\
      8:10& 0.00213633& 1.019811&  0.00955894&     0.198627\\
     10:15& 0.00210643& 2.097187&  0.00945586&     0.398312\\
     20:20& 0.00205055& 0.347043&  0.00931157&     0.074799\\
 \hline
\end{tabular}
\end{center}

@@

\subsubsection{A comparision of Multinomial Naive Bayes to Logistic Regression with Extremely Sparse Feature Matrices}
\noindent
Under extreme levels of sparseness (sparseness measure of less than $0.0021$) from feature matrices generated by CountVectorizer, Logistic Regression and Logistic Lasso break down with greater sparseness while Multinomial Naive Bayes can still produce usable results. \vertSpace

However, we found with less matrix sparsity and a lower Logistic Lasso model penalty (i.e., $\alpha=0.002$), the Logistic Lasso model performance improves.  Based on F1 scores we observed that Logistic Lasso with low penalties performs as well or better than Naive Bayes when sparseness exceeded $0.0025$.  \vertSpace

Logistic Lasso with a larger penalty $>2.0$ does not perform will with extremely sparse matricies. \\
\begin{center}
\noindent \begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline

@@

 \hline
 \multicolumn{1}{|c|}{N-gram} &
 \multicolumn{1}{|c|}{sparsity} &
 \multicolumn{1}{|c|}{smooth$=0.0$} &
 \multicolumn{1}{|c|}{$\alpha=0.002$} &
 \multicolumn{1}{|c|}{$\alpha=1.000$} \\
 \hline
 60& 0.001979 & 0.647 & 0.463 & 0.256\\
 50& 0.002018 & 0.619 & 0.565 & 0.342\\
 30& 0.002038 & 0.724 & 0.620 & 0.460\\
 25& 0.002063 & 0.720 & 0.664 & 0.490\\
 20& 0.002067 & 0.725 & 0.678 & 0.520\\
 12& 0.002069 & 0.682 & 0.743 & 0.656\\
 \hline
\end{tabular}
\end{center}

\begin{figure}[h!]
  \includegraphics[width=\linewidth]{../output/analysis/unstacked-nb-ll.JPG}
  \caption{F1 performance of LL and NB - Unstacked Model}
\end{figure}

\section{Conclusion}
Reducing the sparsity of the matricies generated from CountVectorizer by stacking the documents greatly improved the results.  \vertSpace

We were able to attain maximum test accuracy of $95.9\%$ using a $10:15$ n-gram, producing a recall of $93.4\%$ and an F1 score of $92.5\%$.  Because of the large number (i.e., millions) of features in the feature matrix, Logistic Lasso with small penatly proved to be the most effective method for reducing the number of coefficients, leading to improved F1 scores.  \vertSpace

The fraction of non-zero coefficients remaining after applying Logistic Lasso ranged between $3e-05$ to $1.6e-03$.  Logistic Lasso was very effective in dealing with feature matrix sparsity, and based on what we learned from HW1 question two is the appropriate model choice for our classification problem.  In order to illustrate the effectiveness, we measured the fraction of non-zero coefficients using a Logistic Regression model and the feature matrix had $\approx 97\%$ non-zero coefficients.  \vertSpace

Logistic Lasso (and Logistic Regression generally) broke down at the most extreme levels of sparsity $<0.0020$.  In general, we found that setting the regularization parameters at the extremes ($\alpha=0.0$ or $\alpha > 2.0$) produced inferior results.  \vertSpace

Whereas Multinomial Naive Bayes with Laplace smoothing $=0$ was the most stable method for producing useful results at the most extreme measures of matrix sparcity.  In general we found that when using Multinomial Naive Bayes setting the parameters at no smoothing or large smoothing proved to give the best results. \vertSpace

Overall our research found that Logistic Lasso with $\alpha = 0.002$ produced the most stable results for predicting FOMC rate decisions from FOMC statements, minutes, and Board of Governors speeches.
